Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-19 latex/intro.tex
add explanation for why clusterization

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

@@

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov2012}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

In spatial problems involving choice of measurements within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. This approach incurs a significant
computational cost as it requires solving a difficult combinatorial
optimization problem for measurement locations over a finite set. The
combinatorial optimization problem is usually relaxed by first
assigning optimal measurement weights in $\mathbb{R}_+$ to the
potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega$ fundamentally changes the optimal design
problem and thus results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, researchers recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, they strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

@@

\begin{enumerate}
\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:vanishing}, we provide an insightful
  explanation for the optimality of clustered designs when no model
  error is present. We demonstrate that measurement clusterization
  poses no obstruction to D-optimality and that clustered designs are
  preferred by numerical algorithms.

@@

\item \label{q:avoid} \textbf{Should we aim to avoid measurement
  clusterization?} Based on the analysis conducted in this study, we
  did not find any compelling reason to explicitly avoid optimal
  clustered designs.

\item \label{q:replace} \textbf{Is it possible to substitute an
  optimal clustered design with an equally optimal non-clustered
  design?} In Section \ref{section:vanishing}, we answer this question
  in the affirmative, although we show that numerical experiments
  conducted using our model indicate a strong preference for clustered
  designs.
\end{enumerate}

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties across all posterior
covariance eigenvectors. A similar conclusion was reached by Koval et
al.~\cite{koval2020}, who showed that A-optimal designs are best
constructed in the space of observations.

We first give some general definitions, and then state Theorem
\ref{thm:char}: Let $\hilp, \hilo$ Hilbert spaces, $\fwd:\hilp \to
\hilo$ a linear compact operator. Let $\obs: \hilo \to \mathbb{R}^m$ a
linear measurement operator, where $m \in \mathbb{N}$ is the number of
measurements taken. Let $\sigma^2 \in \mathbb{R}_{+}$ observation
noise variance, $\data = \obs \fwd \param + \eps$, where $\eps \in
\mathbb{R}^m$ is iid $\mathcal{N}(0, \sigma^2)$ noise. Let $\pr \sim
\mathcal{N}(\prmean, \prcov)$ prior Gaussian measure on $\hilp$, where
$\prcov$ is the prior covariance operator and let $\post$ the
posterior (Gaussian) measure.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue it is too far removed
from any real application. To these claims I would argue that
scientists have a long history of studying models that are bare-bones
simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.

>>>>

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

@@

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov1997, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov1997}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

In spatial problems involving choice of measurements within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2022, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. This approach incurs a significant
computational cost as it requires solving a difficult combinatorial
optimization problem for measurement locations over a finite set. The
combinatorial optimization problem is usually relaxed by first
assigning optimal measurement weights in $\mathbb{R}_+$ to the
potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega$ fundamentally changes the optimal design
problem and thus results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, researchers recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, they strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Is measure clusterization a generic phenomenon?
%
And, most importantly: Why does measurement clusterization occur?
%
%Should we aim to avoid measurement clusterization?
%
%Is it possible to substitute an optimal clustered design with an
%equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

@@

\begin{enumerate}

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates $\obs$
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect that measurement clusterization to be a generic phenomenon.

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:why}, we present a compelling
  explanation for the optimality of clustered designs in the absence
  of model error. Our analysis reveals that, in our model, a D-optimal
  design focuses on a select set of prior eigenvectors, specifically
  those with the highest power spectrum. In practical scenarios, the
  number of locations where the relevant prior eigenvectors are
  significantly large, while other eigenvectors have a power spectrum
  of zero, is limited. Consequently, the clustering of measurements
  arises as a natural consequence of the pigeonhole principle, as
  there are more measurements available than there are locations
  satisfying conditions (a) and (b).

  %% In Section \ref{section:vanishing}, we provide an insightful
  %% explanation for the optimality of clustered designs when no model
  %% error is present. We demonstrate that for our model, a D-optimal
  %% design measures only a small subset of prior eigenvectors, which are
  %% the prior eigenvectors with largest power spectrum. In real-life
  %% problems there are limited number of locations where: (a) the
  %% relevant prior eigenvectors are large, and other prior eigenvectors
  %% with large power in the prior spectrum are zero. Then measurement
  %% clusterization is a result of the pigeonhole principle, where there
  %% are more measurements than measurement locations satisfying (a) and
  %% (b).

@@

%% \item \label{q:avoid} \textbf{Should we aim to avoid measurement
%%   clusterization?} Based on the analysis conducted in this study, we
%%   did not find any compelling reason to explicitly avoid optimal
%%   clustered designs.

%% \item \label{q:replace} \textbf{Is it possible to substitute an
%%   optimal clustered design with an equally optimal non-clustered
%%   design?} In Section \ref{section:vanishing}, we answer this question
%%   in the affirmative, although we show that numerical experiments
%%   conducted using our model indicate a strong preference for clustered
%%   designs.

\end{enumerate}

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties for posterior
covariance eigenvectors in observation space. A similar conclusion was
reached by Koval et al.~\cite{koval2020}, who showed that A-optimal
designs are best constructed in the space of observations.

We first give some general definitions before we state Theorem
\ref{thm:char}: Let $\hilp, \hilo$ Hilbert spaces, $\fwd:\hilp \to
\hilo$ a linear compact operator. Let $\obs: \hilo \to \mathbb{R}^m$ a
linear measurement operator, where $m \in \mathbb{N}$ is the number of
measurements taken. Let $\sigma^2 \in \mathbb{R}_{+}$ observation
noise variance, $\data = \obs \fwd \param + \eps$, where $\eps \in
\mathbb{R}^m$ is iid $\mathcal{N}(0, \sigma^2)$ noise. Let $\pr \sim
\mathcal{N}(\prmean, \prcov)$ prior Gaussian measure on $\hilp$, where
$\prcov$ is the prior covariance operator and let $\post$ the
posterior (Gaussian) measure.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue our model is too far
removed from any real application. To these claims I would answer that
scientists have a long history of studying models that are bare-bones
simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.

### 2023-05-19 latex/paper.tex
add explanation for why clusterization

%% \author{Yair Daon\thanks{Porter School of the Environment and Earth
%%     Sciences, Tel Aviv University, Tel Aviv, Israel
%%     (\email{yair.daon@gmail.com}).}}
\author{Yair Daon\thanks{Azrieli Faculty of Medicine, Bar Ilan
    University, Safed, Israel.  (\email{yair.daon@gmail.com}).}}

@@

\input{intro}
\input{prelims}
\input{constrained}
\input{nonvanishing}
\input{vanishing}

%% \section{Open Problems}

\section{Acknowledgements}
This study is a result of research I started during my PhD studies
under the instruction of Prof. Georg Stadler of the Courant Institute
of Mathematical Sciences. I would like to thank him for his great
mentorship, attention to details and kind behavior. This work was
supported by The Raymond and Beverly Sackler Post-Doctoral
Scholarship.

@@

\bibliographystyle{siamplain}
%% %%\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/library.bib}

>>>>

\author{Yair Daon\thanks{Porter School of the Environment and Earth
    Sciences, Tel Aviv University, Tel Aviv, Israel
    (\email{yair.daon@gmail.com}).}}
%% \author{Yair Daon\thanks{Azrieli Faculty of Medicine, Bar Ilan
%%     University, Safed, Israel.  (\email{yair.daon@gmail.com}).}}

@@

\input{intro}
\input{prelims}
\input{constrained}
\input{nonvanishing}
\input{vanishing}
\input{why}
%% \input{how}
%% \section{Open Problems}

\section{Acknowledgements}
This study is a result of research I started during my PhD studies
under the instruction of Prof. Georg Stadler at the Courant Institute
of Mathematical Sciences. I would like to thank him for his great
mentorship, attention to details and kindness. This work was also
supported by The Raymond and Beverly Sackler Post-Doctoral
Scholarship.

@@

\bibliographystyle{siamplain}
%% %%\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/bibtex.bib}

### 2023-05-19 latex/prelims.tex
add explanation for why clusterization

%% \subsection{Overview}
In this section we present notations that will be used throughout this
article. These notations are summarized in the definition of Theorem \ref{thm:char}. The theoretical foundations for inverse
problems over function spaces can be found in \cite{Stuart10}.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces,\footnote{The
subscripts p and o are for ``parameter'' and ``observation'',
respectively.} and denote $\hilo^* \simeq \hilo$ the Hilbert space of
all linear functionals on $\hilo$. Let $\fwd: \hilp \to \hilo$ the
\emph{forward operator}. The forward operator $\fwd$ is assumed linear
and strongly smoothing (has fast decaying modes --- the heat operator
of the 1D heat equation described in Section \ref{section:intro} is a
prime example). Take a Gaussian prior $\param \sim \pr =
\normal(\prmean ,\prcov)$ with some appropriate covariance operator
$\prcov$ on $\hilp$ \cite{Stuart10}. Note that $\fwd \prcov \fwd^*$ is
the prior covariance in $\hilo$ \cite{Stuart10}, and as such, $\fwd
\prcov \fwd^*$ is assumed invertible --- an assumption which will be
used below\footnote{If $\fwd$ has a nontrivial kernel we utilize
Occam's Razor and ignore said kernel.}. Let $m$ the number of
measurements taken. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, whose entries $\meas_j, j=1,\dots,m$ are
called \emph{measurements}:

\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}
%
%

%% For $u \in \hilo$ and $v\in \R^m$:
%% %
%% %
%% \begin{align*}
%%   \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
%%   = v^t \left ( \obs u \right ) = (v^t \obs) (u),
%% \end{align*}
%% and thus:
%% \begin{align}\label{eq:obs*}
%%   \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
%% \end{align}

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error $\eps' \sim
\normal(0,\modcov)$ with $\modcov$ a covariance operator. Observation
error is denoted $\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m \in
\mathbb{R}^{m \times m}$ the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}
%% \subsubsection{Error Terms}
It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
%% and $e_j \in \R^m$ is the $j$th standard basis vector.
%% The explicit dependence on $\obs$ will be mostly dropped for
%% convenience, so $\Sigma(\obs) = \Sigma$.  Thus, for a fixed $\obs$
%% (i.e.\ a fixed set of observations), $\data = \obs \fwd \param +
%% \bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$.
Taking $\modcov = 0$ is a common practice
\cite{Tarantola05,Kaipio2006,Vogel02} and then $\Sigma = \sigma^2I_m$
is a scalar matrix which does not depend on $\obs$.

%% Taking an error model with a non-scalar covariance allows us to
%% consider model error (modeled by $\modcov$) as well as observation
%% error (modeled by $\sigma^2$). For example, say we believe our
%% forward model does not capture some small scale phenomenon.  Then
%% we may express this belief by saying $\tru = \fwd + \err$, with
%% $\fwd$ depending on $\param$ and $\err$ depending on $\sspar$, with
%% $\param \perp \sspar$. We do not know much about this effect but it
%% is reasonable to assume it changes continuously in our domain. We
%% (may choose to) model it as $\normal (0, \modcov)$ and take
%% $\modcov$ to reflect the spatial (or other) variability we imagine
%% $\err \sspar$ has. Such small scale phenomenon can arise as a
%% modeling issue, where we might not model the system in its
%% entirety. It can also arise from a numerical source, where our
%% discretization of the system is not fine enough to capture all
%% small scale phenomena.  In section \ref{section:non vanishing}, we
%% will see that assuming some correlation in the error mitigates the
%% clusterization phenomenon, as reported in the literature
%% \cite{Ucinski05}.

%% \subsubsection{Dynamics and Observation
%% Operators}\label{subsec:dynamics} Rows of $\obs$ are linear
%% functionals and we refer to these as {\it observation
%% vector}s. Note that we cannot choose every observation vector we
%% want. For example, we may be restricted only to pointwise
%% evaluations of $\fwd \param$ by the sensors at our disposal, in
%% which case measuring, e.g., the mean $\ell(u) = \int_{\Omega}u$ of
%% a function is not possible. Thus, it is best to think of $(\obs
%% \fwd \param)_j$ as approximations to pointwise evaluations. For
%% example, if $\hilo = L^2([-\pi, \pi])$, consider $\ell_1(x) :=
%% (2\pi \eps)^{-\frac{1}{2}}\exp(-\frac{x^2}{2\eps^2})$ and
%% $\ell_2(x) :=\eps^{-1}\mathbf{1}_{[-\frac{\eps}{2},
%% \frac{\eps}{2}]}$. Both $\ell_1$ and $\ell_2$ approximate a point
%% measurement at $0$ for small $\eps$ and, crucially, $\ell_1, \ell_2
%% \in \hilo^*$.  %% It is important to note that although $\|\ell\|_1
%% = %% \|\ell_2\|_1=1$, unfortunately $\|\ell_1\|_2^2 = (4\pi %%
%% \eps)^{-\frac{1}{2}}$ and $\|\ell_2\|_2^2 = \eps^{-1}$.

It is useful to record that the posterior measure $\post$ is Gaussian
in this setting and its covariance operator is \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

@@

%% , and just like in finite dimensions, a D-optimal design maximizes
%% the Kullback-Leibler divergence between posterior and prior
%% measures For a linear inverse problem with Gaussian prior and
%% Gaussian noise, a D-optimal design minimizes an expression
%% analogous to the determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).  D-optimal designs in infinite-dimensional Hilbert spaces
%% were investigated in \cite{AlexanderianGloorGhattas14}.
\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}

>>>>

In this section we present the setup that will be used throughout this
article. The theoretical foundations for inverse problems over
function spaces can be found in \cite{Stuart10} and will not be
reviewed here.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces,\footnote{The
subscripts p and o are for ``parameter'' and ``observation'',
respectively.} and let $\fwd: \hilp \to \hilo$ the \emph{forward
operator}. The forward operator $\fwd$ is assumed linear and strongly
smoothing (the heat operator of the 1D heat equation introduced in
Section \ref{section:intro} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ on $\hilp$ \cite{Stuart10}. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}, and as such, $\fwd \prcov \fwd^*$ is assumed
invertible --- an assumption which will be used below\footnote{If
$\fwd$ has a nontrivial kernel we utilize Occam's Razor and ignore
said kernel.}. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, where $m$ is the number of measurements
taken. Entries $\meas_j, j=1,\dots,m$ of the observation operator
$\obs$ are called \emph{measurements}:

\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error $\eps' \sim
\normal(0,\modcov)$ with $\modcov$ a covariance operator. Observation
error is denoted $\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m \in
\mathbb{R}^{m \times m}$ the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking $\modcov = 0$ is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then $\Sigma =
\sigma^2I_m$ is a scalar matrix which does not depend on $\obs$.

It is useful to record that the posterior measure $\post$ is Gaussian
in this setting and its covariance operator is \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}
Note that the poserior covariance $\postcov$ does not depend on data
$\data$.

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}
